---
title: "p8105_hw2_smm2350"
author: "Shaina Mackin" 
date: "2021-10-09"
output: github_document
---

# Homework 2

```{r libraries, echo = FALSE, message = FALSE}
library(tidyverse)
library(readxl)
```

## Problem 1

### Read and Clean

First, I will read and clean the Mr. Trash Wheel sheet, including specifying the sheet and omitting non-data entries using `read_excel`, using reasonable variable names, omitting rows without dumpster-specific data, and rounding the number of sports balls to the nearest integer:

```{r trash}
trash_df = 
  read_excel("data files/Trash.xlsx", 
  sheet = "Mr. Trash Wheel", 
  range = "A2:N406",
  col_names = TRUE) %>%
janitor::clean_names() %>%
drop_na() %>%
mutate(sports_balls = round(sports_balls), digits = 0)

view(trash_df)
```

Second, I will read and clean the precipitation data for 2018 and 2019, including omitting rows without precipitation data and adding a variable for year: 

```{r precipitation_2018_2019}
precip2018_df = 
  read_excel("data files/Trash.xlsx",
  sheet = "2018 Precipitation",
  skip = 1,
  col_names = TRUE) %>%
janitor::clean_names() %>%
drop_na() %>%
mutate(year = 2018)

view(precip2018_df)
  
precip2019_df = 
  read_excel("data files/Trash.xlsx",
  sheet = "2019 Precipitation",
  skip = 1,
  col_names = TRUE) %>%
janitor::clean_names() %>%
drop_na() %>%
mutate(year = 2019)

view(precip2019_df)
```

Next, I will combine the two precipitation datasets and convert month to a character variable: 

```{r precipitation_combined}
precip_df = 
  bind_rows(precip2018_df, precip2019_df) %>%
  janitor::clean_names() %>%
  mutate(month = month.name[as.numeric(month)])
view(precip_df)
```

The Mr. Trash Wheel dataset has `r nrow(trash_df)` observations. The median number of sports balls in a dumpster in 2019 was `r trash_df %>% filter(year == 2019) %>% pull(sports_balls) %>% median()` balls. In addition to sports balls, the dataset records number of `cigarette butts`, `glass bottles`, `plastic bottles`, `polystyrene`, `grocery bags`, and `chip bags` every `month` for `r min(pull(trash_df, year))` to `r max(pull(trash_df, year))`. The precipitation dataset has `r nrow(precip_df)` observations, or months. The total precipitation in 2018 was `r precip_df %>% filter(year == 2018) %>% summarise(sum(total))` inches. The precipitation dataset includes the following variables: `r colnames(precip_df)`.

## Problem 2 

I will now clean the data in the `pols` and `snp` datasets.

```{r pols_snp}
pols_df = 
  read_csv("data files/fivethirtyeight_datasets/pols-month.csv") %>%
  janitor::clean_names() %>%
  separate(mon, c("year", "month", "day")) %>%
  mutate(month = month.name[as.numeric(month)]) %>%
  mutate(year = as.numeric(year)) %>%
  mutate(president = factor(prez_dem, labels = c("gop", "dem"))) %>%
  select(-prez_dem, -prez_gop, -day)
view(pols_df)

snp_df = 
  read_csv("data files/fivethirtyeight_datasets/snp.csv") %>%
  janitor::clean_names() %>%
  mutate(date = lubridate::mdy(date)) %>%
  separate(date, c("year", "month", "date")) %>%
  mutate(month = month.name[as.numeric(month)]) %>%
  mutate(year = as.numeric(year)) %>%
  mutate(year = ifelse(year > 2021, year - 100, year)) %>%
  select(-date)
  
view(snp_df)
```

Next, I will tidy the `unemployment` data so that it can be merged with the previous datasets. 

```{r unemployment}

unemployment_df = read_csv("data files/fivethirtyeight_datasets/unemployment.csv")
view(unemployment_df)

unemployment_tidy_df = 
  read_csv("data files/fivethirtyeight_datasets/unemployment.csv") %>%
  pivot_longer(
    Jan:Dec,
    names_to = "month",
    values_to = "unemployment") %>%
  mutate(month = match(month, month.abb)) %>%
  mutate(month = month.name[as.numeric(month)]) %>%
  janitor::clean_names() 
view(unemployment_tidy_df)
```

Lastly, I will join the datasets by first merging `snp` and `pols` and then merging `unemployment` into the result. 

```{r}
snp_pols_df = 
  left_join(pols_df, snp_df, by = c("year", "month"))

snp_pols_unemploy_df = 
  left_join(snp_pols_df, unemployment_tidy_df, by = c("year", "month"))
```

The `pols` dataset contained `r nrow(pols_df)` observations, or dates, ranging from `r min(pull(pols_df, year))` to `r max(pull(pols_df, year))`, with `r ncol(pols_df)` variables all related to the number of `gop` or `dem` politicians in office at that time. The `snp` dataset contained `r nrow(snp_df)` observations, or dates, ranging from `r min(pull(snp_df, year))` to `r max(pull(snp_df, year))`, with the variable `close` referring to the corresponding closing value of the S&P stock index at that time. The `unemployment` dataset originally contained `r nrow(unemployment_df)` observations, or years, with `month` variables and their corresponding percentages of unemployment. The tidied `unemployment` dataset contains `r nrow(unemployment_tidy_df)` observations, or dates, ranging from `r min(pull(unemployment_tidy_df, year))` to `r max(pull(unemployment_tidy_df, year))`, with their corresponding `unemployment` percentages. The joined `snp_pols_unemploy` dataset now contains `r nrow(snp_pols_unemploy_df)` observations, or dates, ranging from `r min(pull(snp_pols_unemploy_df, year))` to `r max(pull(snp_pols_unemploy_df, year))`, with `r ncol(snp_pols_unemploy_df)` variables including `close` value, `unemployment` percentage, and political party, whether `dem` or `rep`, of politicians accordingly. 
  